\section{Translation Rules and Semantics Lifting}\label{sec:tr}

In this section, we discuss translation rules and the semantics lifting algorithm.
We assume that there exists a sufficiently powerful extended host language (\STLC{} with extensions), and a DSL will be defined by translation rules to this host language.
We first give informal and formal definitions of translation rules and make explicit some requirements.
Then, we present the semantics lifting algorithm formally with examples.
We illustrate two basic properties of the algorithm: correctness and abstraction.
In the proofs, why structural rules are important will be explained.
After that, the nonstructural parts, i.e. lambda abstraction and application, are discussed.
Finally, we provide some extensions to further refine the algorithm and proofs.

\subsection{Translation Rules}

% After defining a fully featured host language,
%  translation rules are designed to specialize the language constructs of the DSL.
% The difference between a translation rule and a macro is that it is not provided by the host language, 
%  but is an abstraction outside the language.
% The distinction between translation rules and syntactic sugar is that it not only simplifies the syntax, 
%  but also can be used to describe more operations.

A translation rule $tr$ defines a new constructor $c_n$, with shape 
\[ c_n~\mexp_{x_1}\cdots \mexp_{x_n} => \mexp \]
 where free expression variables in $\mexp$ must appear in LHS.
We use $LHS(tr)$ and $RHS(tr)$ to denote the expressions on either sides of $=>$.
We call constructors defined by translation rules \textit{surface constructors}.
% Each constructor has only one signature,
Pattern-based translation rules or those with parameter lists are not allowed.

\begin{requirement}\label{req:tr-unique}
  Each surface constructor must be defined by a unique translation rule.
\end{requirement}

If a closed expression $\cexp$ is constructed by $c_n$,
 then there exists an environment $Σ$ satisfying $Σ(LHS(tr))=\cexp$,
 and $Σ(RHS(tr))$ is called \textit{one-step translation} (or desugaring), 
 written as $\ds{\cexp}$.
\textit{Full translation} is also pretty useful,
 which translates expressions of surface constructors recursively.
But if a rule translates surface constructor $c_n$ to itself, 
 i.e., a recursive translation rule,
 the full translation is not terminating.
Therefore, we impose the following requirement.

\begin{requirement}\label{req:no-recursion}
For a translation rule $tr$, constructors in $RHS(tr)$ must be those of host language, or surface constructors defined earlier.
\end{requirement}

Formally, full translation of a closed expression $\DS{\cexp}$ is defined as:
\begin{align*}
  \DS{\mexp_b} & = \mexp_b \\
  \DS{c_n~\cexp_1\cdots \cexp_n} & = \DS{\ds{c_n~\cexp_1\cdots \cexp_n}} & \text{if $c_n$ is a surface constructor} \\
  \DS{c_h~\cexp_1\cdots \cexp_n} & = c_h~\DS{\cexp_1}\cdots\DS{\cexp_n} & \text{if $c_h$ is a host constructor} 
\end{align*}

For example, given an expression $\<true>~\<and>~(\<false>~\<or>~\<true>)$ of \textsc{Bool}, then:
\begin{align*}
  \ds{\<true>~\<and>~(\<false>~\<or>~\<true>)} & = 
    \<if>~\<true>~(\<false>~\<or>~\<true>)~\<false> \\
  \DS{\<true>~\<and>~(\<false>~\<or>~\<true>)} & = 
    \<if>~\<true>~(\<if>~\<false>~\<true>~\<true>)~\<false>
\end{align*}

The full translation can be naturally extended to bones by fully translating all the expressions recursively.
We use $\DB{b}$ to express the full translation on a bone and omit its formal definition.

In addition, we assume that all the filters are commutable with the full translation.
The filters we have used all satisfy this assumption.

\begin{assumption}\label{asm:fun-ds}
  \[
    F(\mexp_1\cdots \mexp_n)\rr{X} v \quad\miff\quad F(\DS{\mexp_1}\cdots\DS{\mexp_n})\rr{X} \DS{v} \\
  \]
\end{assumption}

Another assumption, requires that translations do not affect the success of pattern matching.
This is a requirement for rules in the host language.
For example, $\Let{λx\!:\!t.~\<if>~e_1~e_2~\<false>}{\HH{e}} \cdots$ is not allowed.
Because $λx\!:\!t.~e_1~\<and>~e_2$ cannot match the above pattern.
It is natural that pattern matching should only rely on the value of evaluation and not on the structure of expression itself.

\begin{assumption}\label{asm:pat-ok}
  Suppose $pat$ is a pattern used in a rule of language. Given $v$:
  \begin{align*}
    Σ=match(v,pat) \quad & \miff\quad Σ'=match(\DS{v},pat) \\[-2pt]
                         & \text{and }\quad \forall \mexp_x \in dom(Σ),\DS{Σ(\mexp_x)}=Σ'(\mexp_x) \text{ and vice versa} \\
    \text{FAIL}=match(v,pat) \quad & \miff\quad \text{FAIL}=match(\DS{v},pat)
  \end{align*}
\end{assumption}

\paragraph{Variable Scope in Translation Rules}

Pombrio et al. \cite{infer-scope} provide a method to infer scope though syntactic sugar.
They add restrictions to the syntactic sugar to keep the scope safe.
We follow the conclusion:

\begin{requirement}\label{req:close}
  A translation rule must be \textit{closed}: 
    any variable used in the RHS must be bound locally (by lambda abstraction or let binding).
\end{requirement}

The requirement prohibits external variables capture.
Also, locally bound variables must be explicitly parameterized for use in other parameters.

\begin{example}
The definition of two translation rules and their sample programs are shown below:
\[
  \begin{array}{cl|cl}
    % \text{Translation Rules} & & \text{Sample Programs} \\
    \mathit{leaked}~e => \<let>~x:\<int> =1~\<in>~e
    & \text{(\ding{51})}
    & \mathit{leaked}~(x+1)
    & \text{(\ding{55})} \\
    \mathit{captured}~e => \<if>~x~\<true>~e
    & \text{(\ding{55})}
    & \<let>~x:\<bool> =\<true>~\<in>~\mathit{captured}~\<false>
    & \text{(\ding{55})}
  \end{array}
\]

The $\mathit{leaked}$ tries to bind a constant in $x$ for use in $e$.
But $x$ referenced in $e$ is not allowed.
Thus the translation rule is actually premitted,
 but $x$ is not in scope of $e$.
As a comparison, the following definition is compliant:
\[ \mathit{leaked}'~x~e => \<let>~x:\<int> =1~\<in>~e \]
The $\mathit{captured}$ attempts to get the value of $x$ in the current environment,
 which may produce unbound identifiers after translation.
The translation rule itself is not allowed.
\end{example}

% Taking all these considerations into account, we give the following requirements.

\subsubsection{Hygiene}

Programming languages with non-hygienic macro systems may cause the hygiene problem\cite{hygine}:
  variable bindings are possibly hidden by macros, and vice versa.
For example, we define a translation rule $or'$ via $\<let>$:
\[ e_1~or'~e_2 => \<let>~x:\<bool> =e_1~\<in>~\<if>~x~x~e_2 \]
 where $``x"$ is a literal identifier. 
Then the expression $\<let>~x:\<bool> =\<false>~\<in>~(\<true>~or'~x)$ will be fully translated into $\<let>~x=\<false>~\<in>~\<let>~x=\<true>~\<in>~\<if>~x~x~x$, causing an error.

The closed translation rules ensure no unbound identifier exceptions \cite{infer-scope}.
Therefore, we can modify the names of variables safely just like $\alpha$--equivalence.

\begin{assumption}
  We treat literals variables bound in the RHS as mutable and always fresh.
  And the translation is always hygienic.
\end{assumption}

% We use $\texttt{@}$ to denote fresh variables, and $or'$ will be written as:
% \[ e_1~or'~e_2 => \<let>~\texttt{@x}=e_1~\<in>~\<if>~\texttt{@x}~\texttt{@x}~e_2. \]

% \subsubsection{Substitution Rules}

For each language construct defined by a translation rule,
 the developer needs to define substitution rules for it,
 just like those of the host language.
 In fact, the substitution rule should be automated by the derivation of the scope.
% The set of substitution rules in DSL $R'_{subst}$ is a union of the substitution rules of the host language $R_{subst}$ and these newly defined substitution rules.
% Since substitution is a pure meta-function, the assumption \ref{asm:fun-ds} should be satisfied in DSL.
% Intuitively, substitution and full translation are commutable.
% Deriving substitution rules automatically is our future work.

\subsection{Semantics Lifting}

% In order for the DSL to support new language constructs defined by translation rules, 
%  it is necessary to provide their evaluation rules and type rules.
% In this section, a generic algorithm will be presented, 
%  to illustrate how to derive the rules for translation rules.
% For a DSL defined by multiple translation rules, 
%  we derive the evaluation and typing rules and add them to the language individually.

% And now we will give the algorithm of semantics derivation. 
Suppose a host language $L$ based on \STLC{}. % is defined as $L=\langle S_p, C, R_{\mE}, R_{\mT}, R_{subst}, F\rangle$,
%  where $S_p=\{\Exp,\Type\}$.
We now have a set of translation rules to define DSL.
Due to requirement \ref{req:no-recursion},
 we can simplify the problem to process one translation rule one by one.
% And the language which supports the new language constructs is used as a new host language,
%  to derive evaluation rules for other translation rules.
Suppose the translation rule $tr$ has shape $c_n~exp_1\cdots exp_n => e$,
 and $sort(c_n)=\Exp$.
% Our goal is to get the definition of new language $L'=\langle S_p, C', R'_{\mE}, R'_{\mT}, R'_{subst}, F\rangle$,
%  which supports the new language construct $c_n$.
% The definitions of $C'$ and $R_{subst}$ are obvious.
% We need to define the evaluation and typing rule for $c_n$.
% By the method proposed by \cite{infer-types}, the typing rule is obtained.
% In the following, we will talk about how to derive the evaluation rule.
We aim to get the evaluation rule of $c_n$.

The core idea of the semantics derivation has been shown in overview:
 expand evaluation of compound expressions recursively according to the rules in $R_{\mE}$
 until all the evaluation are unexpandable.
Formally, $\ctr{Name}{LHS(tr)}=\dd(\HH{RSH(tr)})$, the definition of $\dd$ is shown in Fig. \ref{fig:sd}.
Note that here we skip the transformation of expressions containing substitutions for now.

\begin{figure}
  $\ctr{Name}{c_n~exp_1\cdots exp_n}=\dd(\HH{e})$
  \begin{align}
    \noalign{\raggedright \hspace{2em} \bf Bone:}
    \dd(b:(br_1\cdots br_n)) & = \dd(b):(\dd(br_1)\cdots \dd(br_n)) \\
    \dd(\Let{pat}{b_1} b_2) & = \Let{pat}{\dd(b_1)} \dd(b_2) \\
    \dd(\HH{\mexp_x}) & = \HH{\mexp_x} \\  
    \dd(\HH{\mexp_b}) & = \HH{\mexp_b} \\
    \hspace{8em} \dd(\HH{c_h~\mexp_1\cdots \mexp_m}) 
      & = \dd(Σ(b)) \hspace{20pt} \text{if there exists $Σ$ and $\ctr{C}{\mexp}\cqq b$ in $L$,} \\
      & \hspace{61pt} \text{such that $Σ(\mexp)=c_h~\mexp_1\cdots \mexp_m$} \nonumber \\
    \dd(\HH{\mexp[\mexp'/x]}) & = \HH{\mexp[\mexp'/x]} \\
    % \dd(\EE{f_p(\mexp_1'\cdots \mexp_n')}) & = \EE{f_p(\mexp_1'\cdots \mexp_n')} \\
    \dd(F(\mexp_1\cdots \mexp_n)) & = F(\mexp_1\cdots \mexp_n) \\
    \noalign{\raggedright \hspace{2em} \bf Expression:}
    \dd(\mexp) & = \mexp \\
    \noalign{\raggedright \hspace{2em} \bf Branch:}
    \dd(pat |> b) & = pat |> \dd(b)
    % \dd(pat \mid \mexp' |> b) & = pat \mid \mexp' |> \dd(b)
  \end{align}
  \caption{Semantic Derivation}
  \label{fig:sd}
\end{figure}

To illustrate the algorithm, some examples will be presented.
Example \ref{exm:nand} defines $\<nand>$ by $\<not>$ and $\<and>$.
Assume that the evaluation rules of $\<not>$ and $\<and>$ have been derived earlier.
We will observe that how $\dd$ works recursively.
For clarity, the parts applied by $\dd$ are underlined.

\begin{example}\label{exm:nand}
  $e_1~\<nand>~e_2 => \<not>~(e_1~\<and>~e_2)$:
  \begin{align*}
       & \ctr{Nand}{e_1~\<nand>~e_2} \\
    =~ & \dhl{\HH{\<not>~(e_1~\<and>~e_2)}} \\
    =~ & \dhl{\HH{e_1~\<and>~e_2}
            :\branch{\<true>|>\<false> \\& \<false>|>\<true>}} \\
    =~ & \dhl{\HH{e_1~\<and>~e_2}}
            :\branch{\<true>|>\dhl{\<false>} \\& \<false>|>\dhl{\<true>}} \\
    =~ & \dhl{\HH{e_1}
            :\branch{\<true>|>\HH{e_2} \\& \<false>|>\<false>}
           }:\branch{\<true>|>\<false> \\& \<false>|>\<true>} \\
    =~ & \HH{e_1}
          :\branch{\<true>|>\HH{e_2} \\& \<false>|>\<false>}
          :\branch{\<true>|>\<false> \\& \<false>|>\<true>} 
  \end{align*}  
\end{example}

We pick up the $\<let>$ defined in introduction as the second example.
% The language construct $\<let>$ is a common syntactic sugar, defined by lambda abstraction and application.
% Considering $\<let>$ as a translation rule, 
Example \ref{exm:let} presents the derivation of its evaluation rule.
% Note that the RHS of the translation rule satisfies \ref{ll:let}, and no language constructs will be generated in lambda lifting.
The last step is simplification, since that both left and right sides are constructed by $λ$.
Simplification is sometimes useful for abstraction property, for instance in this example,
 where the evaluation rule of $\<let>$ no longer depends on lambda abstraction (even if this is allowed).
% In addition, substituion, as a meta-function, appears in the following derivation, which does not affect the expansion.
% In \todo{}, we will discuss substituion more deeply.

\begin{example}
  $\<let>~x:t=e_1~\<in>~e_2 => (λx:t.e_2)~e_1$:
  \begin{align*}
       & \ctr{Let}{\<let>~x:t=e_1~\<in>~e_2} \\
    =~ & \dhl{\HH{(λx:t.e_2)~e_1}}  \\
    =~ & \dhl{\Let{λx':t'.e}{\HH{λx:t.e_2}} \Let{v_1}{\HH{e_1}} \HH{e[v_1/x']}} \\
    =~ & \Let{λx':t'.e}{\dhl{\HH{λx:t.e_2}}} \dhl{\Let{v_1}{\HH{e_1}} \EE{e[v_1/x']}} \\
    =~ & \Let{λx':t'.e}{λx:t.e_2} \Let{v_1}{\dhl{\HH{e_1}}} \dhl{\HH{e[v_1/x']}} \\
    =~ & \Let{λx':t'.e}{λx:t.e_2} \Let{v_1}{\HH{e_1}} \HH{e[v_1/x']} \\
    =~ & \Let{x'}{x} \Let{t'}{t} \Let{e}{e_2} \Let{v_1}{\HH{e_1}} \HH{e[v_1/x']} \\
    =~ & \Let{v_1}{\HH{e_1}} \HH{e_2[v_1/x]}
  \end{align*}
  \label{exm:let}
\end{example}

\subsection{Properties}

With the semantics lifting algorithm, the evaluation rule of $c_n$ is generated.
We expect the rule to have the following properties:
\begin{itemize}
  \item Correctness: a closed expression $\ce$ in $L'$, should have the same value as 
         that of expression fully translated first and then evaluated in $L$.
  \item Abstraction: any constructor used in evaluation rule of $c_n$ is either a constructor of values,
         or a DSL constructor.
\end{itemize}
In this section, we will presents these properties formally.

\subsubsection{Correctness}
We use subscripts to distinguish between in which language the computation is performed.
Then, the correctness is
\[ \EELL{\ce} \rr{X} v \quad\miff\quad \EEL{\DS{\ce}} \rr{X} \DS{v}. \]

% For the proof of the correctness theorem,
%  we first illustrate that $\dd$ and environment application is commutable,
%  and $\dd$ does not influence the result of a closed bone $\cb$.

For the proof of the correctness theorem,
 the following two lemmas are necessary,
 which describe some properties of $\dd$ in the above algorithm.
The complete proof can be found in appendix.

\begin{lemma}\label{lemma:del-sig}
  Support $Σ$ is closed for bone $b$. Then:
  \[ Σ(\dd(b)) \rr{X} v \quad\miff\quad \dd(Σ(b)) \rr{X} v. \]
\end{lemma}

\begin{lemma}\label{lemma:sig-bone}
  Support $\cb$ is a closed bone. Then:
  \[ \cb \rr{X} v \quad\miff\quad \dd(\cb) \rr{X} v. \]
\end{lemma}

Before the proof of correctness, we will give a stronger lemma,
 of which correctness theorem is an special case.

\begin{lemma}\label{lemma:b-db}
  Support $\cb$ is a closed bone. Then:
  \[ \cb \rr{X} v \quad\miff\quad \DB{\cb} \rr{X} \DS{v} \]
\end{lemma}

\begin{proof}
  By induction on a derivation of $\cb \rr{X} v$. 
  The \textsc{Expression} case are obvious.

  Case \textsc{Filter}: by assumption \ref{asm:fun-ds}. 
  
  Case \textsc{Hook}: suppose that $b=\EE{e}$. There are two cases should be considered.
  \begin{itemize}    
    \item $e=c_n~\cexp_1\cdots \cexp_n$ where $c_n$ is a surface constructor defined by $c_n~\mexp_{x_1}\cdots \mexp_{x_n}=>\mexp$.
      Then $\EE{c_n~\cexp_1\cdots \cexp_n} \rr{X} v$ iff $L'[c_n~\mexp_1\cdots \mexp_n]\rr{X} v$.
      According to the definition, there exists an environment $Σ$, such that $Σ=\{ \mexp_{x_i} \mapsto \mexp_i, i\in [1..n] \}$ and $L'[c_n~\cexp_1\cdots \cexp_n]=Σ(\dd(\EE{\mexp}))$.
      Therefore:
      \begin{align*}
        & \EE{c_n~\cexp_1\cdots \cexp_n} \rr{X} v \\
        \miff \quad & L'[c_n~\cexp_1\cdots \cexp_n]\rr{X} v & \text{by the rule \textsc{Rec}} \\
        \miff \quad & \DB{L'[c_n~\cexp_1\cdots \cexp_n]} \rr{X} \DS{v} & \text{inductive hypothesis} \\
        \miff \quad & \DB{Σ(\dd(\EE{\mexp}))} \rr{X} \DS{v}
      \end{align*}

      Because $\mexp$ is an expression defined by host language constructs,
       surface construct $c_n$ will not be used in $\dd(\EE{\mexp})$.
      Let $Σ'=\{ \mexp_{x_i} \mapsto \DS{\mexp_i}, i\in [1..n] \}$.
      Thus, $\DB{Σ(\dd(\EE{\mexp}))} = Σ'(\dd(\EE{\mexp}))$. Then:

      \begin{align*}
        & Σ'(\dd(\EE{\mexp})) \rr{X} \DS{v} \\
        \miff \quad & \dd(Σ'(\EE{\mexp})) \rr{X} v & \text{by lemma \ref{lemma:del-sig}} \\
        \miff \quad & Σ'(\EE{\mexp})\rr{X} v & \text{by lemma \ref{lemma:sig-bone}} \\
        \miff \quad & \EE{Σ'(\mexp)}\rr{X} v
      \end{align*}

      Also:
      \[ Σ'(\EE{\mexp}) = \EE{Σ'(\mexp)} = \EE{\DS{Σ(\mexp)}} = \DS{\EE{Σ(\mexp)}} = \DB{\EE{c_n~\cexp_1\cdots \cexp_n}} \] 
      Thus $\EE{c_n~\cexp_1\cdots \cexp_n} \rr{X} v$ iff $\DB{\EE{c_n~\cexp_1\cdots \cexp_n}} \rr{X} \DS{v}$.
    
    \item $e=c_h~\mexp_1\cdots \mexp_n$ where $c_h$ is a host constructor. Similar.
    % \item $e=f_p(\mexp_1\cdots \mexp_n)$: proved by assumption \ref{asm:fun-ds}.
  \end{itemize}

  Case \textsc{LetIn}: let $b=\Let{pat}{b_1} b_2$.
  Then $b\rr{X} v$, iff the following judgements are satisfied:  
  \[ b_1\rr{X_1}v_1 \quad Σ=match(pat,v_1) \quad Σ(b_2)\rr{X_2}v_2 \quad X=X_1;X_2, \]
  By the induction hypothesis:
  \begin{align*}
    b_1\rr{X_1}v_1    & \quad\miff\quad \DB{b_1}\rr{X_1} \DS{v_1}, \\
    Σ(b_2)\rr{X_2}v_2 & \quad\miff\quad \DB{Σ(b_2)} \rr{X_2} \DS{v_2}
  \end{align*}

  We aim to show that $Σ'=match(pat, \DS{v_1})$ and $Σ'(\DB{b_2})=\DB{Σ(b_2)}$.
  Requirement \ref{req:pat} and assumption \ref{asm:pat-ok} are applied.
  % $v_1$ and $\DS{v_1}$ are either base expressions (trivial case) or results of recursive computation.

  Case \textsc{Branch}: Similar.
\end{proof}

\begin{theorem}[Correctness]
  For any closed expression $\ce$ with sort $\Exp$ in $L'$, 
  \[ \EELL{\ce} \rr{X} v \quad\miff\quad \EEL{\DS{\ce}} \rr{X} \DS{v} \]
\end{theorem}

\begin{proof}
  % For any closed expression $\ce_h$ with sort $\Exp$ in host language $L$,
  % $\EEL{\ce_h} \rr{X} \DS{v}$ iff $\EELL{\ce_h} \rr{X} \DS{v}$. 
  Proved in lemma \ref{lemma:b-db}.
\end{proof}

\subsubsection{Abstraction}

Correctness is the foundation of the algorithm,
 while abstraction is the goal of our algorithm.
Before the proof of abstraction property,
 we first present \textit{strong abstraction theorem}.

\begin{theorem}[Strong Abstraction]\label{thm:str-abs}
  If all the evaluation rules of the host language $L$ are structural,
  the derived rule maintains structural.
\end{theorem}

If the derived rule is structural, it means that all the constructs in it are used to construct values. Therefore, abstractness obviously holds.
Consider the expressions evaluated during the process of $\dd$.
The only interesting case is that the expression is a compound expression constructed by $c$.
Since the evaluation rule of $c$ is structural, the expressions evaluated must be the arguments of $c$.
Due to the finite depth of RHS of the translation rule,
  they eventually converge to the evaluation of parameters,
  or to the (constant) expressions as values.
Because all the rules in expansion are structural, all the constructors are used to construct values.
Therefore, the derived rule is structural.

The theorem \ref{thm:str-abs} can be formalized as follows:
If a translation rule is of the form $c_n~\mexp_{x_1}\cdots\mexp_{x_n}=>\mexp$,
  then $check(\{\mexp_{x_1}\cdots\mexp_{x_n}\}, \dd(H(\mexp)))=\<true>$ holds.

\begin{proof}
  We just need to prove that given a set of expression variable $V$, 
  and an expression $\mexp$ which is constructed by language constructs in $L$ and variables in $V$,
  \[ check(V, \dd(H(\mexp))=\<true> \]

Induct on $\mexp$.
If $\mexp$ is an expression variable in $V$, $\dd(H(\mexp))=H(\mexp)$. The conclusion holds.

If $\mexp$ is a compound expression $c~\mexp_1\cdots\mexp_m$, then there exists a rule of $c$ in language $L$ $\ctr{Name}{c~\mexp_{y_1}\cdots\mexp_{y_m}} \cqq b$ and an environment $\Sigma$ satisfying
$\Sigma=\{\mexp_{y_1}\mapsto \mexp_1,\cdots, \mexp_{y_m}\mapsto \mexp_m\}$, such that $\dd(H(\mexp))=\dd(\Sigma(b))$, and $check(\{\mexp_{y_1}\cdots\mexp_{y_m}\},b)=\<true>$.
Then induct on $b$. % ignoring cases making $check$ false.
Since all the hooks in $b$ is like $H(\mexp_{y_i})$, they will be transformed into $\dd(H(\mexp_i))$.
By inductive hypothesis, $check$ satisfies for $b$.
% (1) If $b$ is of sort $\Val$, without $\mexp_{y_i}$ of sort $\Exp$ used,
% then $\dd(\Sigma(b))=b$.
% (2) If $b=H(\mexp_{y_i})$, $\dd(\Sigma(b))=\dd(H(\mexp_{i}))$.
% By inductive hypothesis, $check(V, \dd(H(\mexp_{i}))=\<true>$ satisfies.
% (3) If $b=\Let{pat}{b_1} b_2$, 
\end{proof}

Strong abstraction property illustrates the importance of structural evaluation rules.
Moreover, new language constructs defined on language constructs with structural evaluation have structural evaluation rules as well.
In STLC, however, not all the language constructs have structural semantics: lambda abstraction and application.

\subsection{Preprocessing: Lambda Lifting}

We have already examplified how lambda abstractions break the abstraction property in semantics lifting.
The solution to this problem is to expose the body of lambda abstractions in the translation rules 
 and define them as new language constructs of the DSL.
For outermost lambda abstractions, the transformation is straightforward. 
Inner lambda abstractions, on the other hand, need to be extracted recursively.
The challenge here is to ensure that the generated translation rules are closed.
Fortunately, lambda lifting can help us.
Formally, for the translation rule $tr$ which defines $c_n'$:
\newcommand{\laml}[1]{\uparrow_λ\!\!(#1)}
\begin{align*}
  \laml{(λx:t.e_1)~e_2} & = (λx:t.\laml{e_1})~\laml{e_2} \label{ll:let} \\
  \laml{λx:t.e} & = λx:t.e \qquad \text{if $e$ is a variable or expression variable} \\
  \laml{λx:t.e} & = λx:t.(c_n'~\mexp_1..\mexp_p ~ x_1..x_q) \\[-3pt]
    & \text{generating } c_n'~\mexp_1..\mexp_p ~ e_1..e_q => \laml{e[e_1/x_1..e_q/x_q]} \\[-3pt]
    & \text{for each expression variable } \mexp_i \text{ in } e \text{ and variable } x_j \text{ not bound in } e \\
  \laml{c~\mexp_1\cdots \mexp_n} & = c~\laml{\mexp_1}\cdots\laml{\mexp_n} \\
  \laml{\mexp_x} & = \mexp_x \\
  \laml{\mexp_b} & = \mexp_b
\end{align*}
where $\laml{\bullet}$ is lambda lifting transformation, with the generation of some translation rules.
The first rule states that if a lambda abstraction is applied directly in the translation rule, 
 then it does not need to be lifted because it does not break the abstraction (which can be written as a $\<let>$).
The second is the main rule, which generates a new constructor $c_n'$ to describe the behaviour of a lambda abstraction.
Any expression variable of $LHS(tr)$ used in $e$ (i.e., parameters of $c_n$) needs to be applied by $c_n'$.
In addition, any variable $x_j$ which has been bound and captured by $e$ is also required to be included,
 by an new expression variable $e_j$, which ensures the closed requirement of translation rules.
Note that the newly generated $c_n'$ needs to be processed by lambda lifting recursively,
 and that may generate some more translation rules.

% The following lemma clearly holds.

\begin{lemma}
  Lambda lifting preserves requirements \ref{req:no-recursion} and \ref{req:close}.
\end{lemma}

The correctness of lambda lifting is pretty obvious.
For abstraction, the inner language constructs are lifted, i.e. replaced by a new DSL construct $c_n'$.

\begin{theorem}
  lambda lifting maintains correctness and abstraction.
\end{theorem}

In addition, Lambda lifting is applicable to any language construct that produces delayed evaluations.

\subsection{Substitution}

In the evaluation rule of application and let binding, substitution is introduced.
As mentioned earlier, we need to specify substitution rules for each language construct.
Except for variable, the substituion rules are all structural.
Some language constructs involving scopes need to check the variable name during substitution.
For example, the substitution rule of lambda abstraction is defined as follows:
\[ (λx:t.~e_1)[e_2/y] = \left\{ \begin{aligned}
  & λx:t.~e_1 & x=y \\
  & λx:t.~e_1[e_2/y] & \text{otherwise}
\end{aligned} \right. \]
And static application of the substitution rule to a evaluation does not result in an error:
\[ \HH{(λx:t.~e_1)[e_2/y]} = \left\{ \begin{aligned}
  & \HH{λx:t.~e_1} & x=y \\
  & \HH{λx:t.~e_1[e_2/y]} & \text{otherwise}
\end{aligned} \right. \]
To describe behaviour with conditions, we simply extend our meta-language with guard (side-condition) in pattern matching.
For example, the above equation can be expressed as:
\[ \HH{(λx\!:\!t.~e_1)[e_2/y]} = ():\branch{
    \_ \mid x=y |> \HH{λx\!:\!t.~e_1} \\&
    \_ |> \HH{λx\!:\!t.~e_1[e_2/y]}} = ():\branch{
      \_ \mid x=y |> λx\!:\!t.~e_1 \\&
      \_ |> λx\!:\!t.~e_1[e_2/y]} \]
where the pattern matching here is to find the satisfied condition.
Therefore, in order to cover the substitution,
 just expand the evaluation containing substitution statically in semantics lifting.
%  The transformation is correct (the CLOSED requirement is important). 
% The abstractness holds since the expressions used for substitution in application are values.
We add the following rule to $\dd$:
\[ \dd(\HH{(c~\mexp_1\cdots\mexp_n)[e/x]}) = ():\branch{
      \_ \mid cond_1 |> \dd(\HH{Σ(\mexp_1')}) \\&
      \cdots \\&
      \_ \mid cond_m |> \dd(\HH{Σ(\mexp_m')}) \\  
} \]
where the substituion rule of $c$ is:
\[ (c~\mexp_{x_1}\cdots\mexp_{x_n})[e/x] = \left\{ \begin{aligned}
      & \mexp_1' & cond_1 \\
      & \cdots \\
      & \mexp_m' & cond_m 
    \end{aligned} \right. \]
and $\mexp_i'$ is expression with substitution, $Σ(c~\mexp_{x_1}\cdots\mexp_{x_n})=c~\mexp_1\cdots\mexp_n$.

\begin{example}
  Consider the translation rule defined:
  \[ e_1~or'~e_2 => \<let>~x:\<bool> =e_1~\<in>~\<if>~x~x~e_2 \]
  The evaluation rule of $\<or>'$ is derived as follows:

  \begin{align*}
    \ctr{OrP}{e_1~or'~e_2} 
    & \cqq \dhl{\HH{\<let>~x:\<bool> =e_1~\<in>~\<if>~x~x~e_2}} \\
    & = \Let{v_1}{\HH{e_1}} \dhl{\HH{(\<if>~x~x~e_2)[v_1/x]}} \\
    & = \Let{v_1}{\HH{e_1}} \dhl{\HH{\<if>~x[v_1/x]~x[v_1/x]~e_2[v_1/x]}} \\
    & = \Let{v_1}{\HH{e_1}} \dhl{\HH{x[v_1/x]}}: \branch{
          \<true> |> \dhl{\HH{x[v_1/x]}} \\&
          \<false> |> \dhl{\HH{e_2[v_1/x]}} } \\
    & = \Let{v_1}{\HH{e_1}} v_1: \branch{
          \<true> |> v_1 \\&
          \<false> |> \HH{e_2[v_1/x]}} 
  \end{align*}

  Since we assume the translation to be hygienic, or rather,
  here $x$ is an arbitrary fresh variable and cannot be used in $e_2$.
  Therefore the rule can be reduced to

  \[ \ctr{OrP}{e_1~or'~e_2} \cqq \Let{v_1}{\HH{e_1}} v_1: \branch{
        \<true> |> v_1 \\&
        \<false> |> \HH{e_2}} \]
\end{example}

Hygienic translation guarantees correctness. 
According to the above rules, the substitution for expressions eventually converges to the substitution for expression variables.
Also, since the expressions used for substitution are values, abstractness holds.

\begin{theorem}
  Substitution maintains correctness and abstraction.
\end{theorem}

\subsection{Summary and Extensions}\label{sec:alg-ex}

We have introduced the semantics lifting algorithm in this section.
We find that translation rules defined with language constructs with structural evaluation rules keeps structural.
The lambda abstraction and application, are specifically discussed so that the semantics lifting algorithm supports, and maintains abstraction.
Thus, for a language containing structural and nonstructural parts,
 the lifting in the structural part is always safe and remains structural.
Therefore, the abstraction is also guaranteed when we define a DSL again based on the current DSL defined by translation rules.
For the nonstructural parts, we have a trade-off that depends on the specific problem as well as the reasonableness of the abstractness breakdown..
Then, we achieve a vertical growth of the language: specialization by translation rules.

Now we are going to extend our algorithm slightly to enhance its usefulness. The use of these extensions is a trade-off.
These extensions may result in abstraction not being guaranteed.

\paragraph{Substitution with non-value expression}

In general, the expressions to substitute are values.
This holds for both application and let binding.
In the above strategy, we require the substitution rule to be applied in the valuation.
However, if the expression used for substitution is not a value,
 such as fixpoint, the above approach may not terminate.
For this case, we add the following rule:
\[ \dd(\HH{e[e'/x]}) = \HH{e[e'/x]} \qquad \text{if $e'$ is not a value}. \]
We discuss fixpoint more extensively in Section \ref{sec:fix}.

\paragraph{Recursive Translation Rules}

Requirement \ref{req:no-recursion} specifies that the translation rule cannot be defined with language construct of itself,
 otherwise the expression cannot be fully translated.
However this is feasible for translation one step at a time. 
This method is called lazy desugaring \cite{lazy-desg}.
Another problem then is how to recursively expand the rules in semantics lifting.
In fact, we cannot find such a solution.
Consider that the new language construct $c_n$ itself is a DSL construct.
We can think of it as satisfying abstraction.
We add the fllowing rule to $\dd$:
\[ \dd(\HH{c_n~\mexp_1\cdots \mexp_n}) = \HH{c_n~\mexp_1\cdots \mexp_n} \]

But only the language construct of its definition itself satisfies the abstraction,
 the expressions as arguments cannot be guaranteed. 
In addition, the translation rules defined by $c_n$ cannot remove the language construct in the derivation,
 and therefore must not satisfy abstraction.
We discuss $\<while>$ as an example in Section \ref{sec:while}.

